Effective Reionization Likelihood from Planck 2018 Data (?)

### 2021-01-07 paper/rei.tex

where we have discretized the redshift space with $\delta z= 0.25$, and where $\sigma_a^2$ are the variances of the PCs. We have rank-ordered the PCs from low to high variance, so that for the range $z_{min} = 6$ (to be consistent with Ly$\alpha$ forest constraints, e.g. \cite{Becker:2015lua}) and $z_{\rm max} = 30$, only the first 5 components are needed to describe all the information on $x_e$ carried by $C_\ell^{EE}$ to cosmic variance limit. In the work that follows, we therefore truncate the sum at $a = 1 .. n_{\rm{PC}}$, where $n_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ and $n_{\rm{PC}} = 7$ for $z_{\rm max} = 50$.

@@

[rephrase this paragraph:]
``For the Planck data set, most of the information in the ionization history is carried by the
first two modes and therefore relates to the amount of high vs.~low redshift optical depth. We keep all 5 PCs  for completeness in representing the observable
impact of a given ionization history and to marginalize uncertainties that they introduce."
We follow Ref.~\cite{Heinrich:2018btc} to compute the CMB power spectra with PCs using a modified version of CAMB\footnote{CAMB: \url{http://camb.info}}~\cite{Lewis:1999bs, Howlett:2012mh}, where the PCs were discretized at $\delta z = 0.25$, and computed at the fiducial model $x_e^{\rm fid} = 0.15$. Note that following Ref.~\cite{Heinrich:2018btc}, we have updated the $\zmax = 30$ PCs to be computed with Planck 2015 rather than WMAP best-fit parameters, which results in minor differences between the PC functions used in Ref.~\cite{Heinrich:2016ojb} and here. For $z<6$, we follow CAMB to assume fully ionized hydrogen and singly ionized helium and for $z\leq 3.5$~\cite{Becker:2010cu}, doubly ionized helium~\cite{Becker:2010cu} with a width $\Delta z = 0.5$. 
\wh{Probably need some brief words about MCMC sampling of the posteriors of $m_a$ jointly with $\Lambda$CDM parameters here so that the KDE section makes sense.  I moved the optical depth stuff to a following subsection to clarify role, and order of operations, of the physicality prior.}

\subsection{Effective Likelihood from PCs -- Kernel Density Estimate}
\label{sec:KDE}
The completeness property of the PCs enables us to turn PC chains obtained from a MCMC run into an effective likelihood that can be used to test any reionization model with the CMB without a cumbersome reanalysis at the level of power spectra data. In the following, we briefly recap the kernel density estimate technique used to build this likelihood and refer the readers to Ref.~\cite{Heinrich:2016ojb} for more details.
The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ for $i = 1$...$N_{\rm sample}$. Given any physical ionization history $x_e(z)$, we first obtain its PC representation $\mathbf{m}$ using Eq.~\ref{eq:xetommu}. Since $\mathbf{m}$ could take any continuous value, we approximate its effective likelihood with a kernel density estimate
\beq
{\cal L}_{\rm PC}\left({\rm data}|\mathbf{m} \right)  = \sum_{i = 1}^{N_{\rm sample}} w_i K_f(\mathbf{m}-\mathbf{m}_i),
\eeq
where the overall normalization is arbitrary, and where we have chosen the smoothing kernel $K_f$ to be a  multivariate Gaussian with mean zero and covariance $f\mathbf{C}$, where $\mathbf{C}$ is the $N_{\rm PC} \times N_{\rm PC}$ covariance matrix estimated from the PC chains (see Table~\ref{tab:PC_stats}) and $f$ is a fraction smaller than 1.
For a Gaussian posterior, increasing the covariance by $1+f$ corresponds to increasing the standard deviation by approximately $1+f/2$. To minimize the amount of smoothing needed while still maintaining good accuracy in the tail of the distribution for any physical models, we oversample the PC distributions by running the PC chains far beyond convergence for about $N_{\sample} \approx 1.1 \times 10^{6}$ chain samples. 
% 1.06 cut w/ physicality prior
% 1.072 no cut
For this $N_{\sample}$, a fraction of $f = 0.14$ should be sufficient. Note also that we employ PC chains without the physicality priors allowing the smoothing kernel to cross over these priors. 
Evaluating the posterior distribution of model parameters $\bf p$ using the effective likelihood
\begin{equation}
P({\bf p}| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\bf p) \right] P(\bf p),
\end{equation}
we demonstrate a successful recovery of the posterior distributions from direct MCMC analyses (using the Planck likelihood and varying all cosmological parameters) with significantly less computational time. We show these results in sections~\ref{sec:example1} and~\ref{sec:example2} for two example models: the standard steplike model and a two-step model allowing for high-redshift ionization.

An even simpler effective likelihood is the Gaussian approximation of the $m_a$ posterior using the PC mean $\bar{\bf m}$ and covariance $\bf C$
%
\begin{equation}
 {\cal L}_{\rm gauss}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^{N_{\rm PC}} | \mathbf{C}| }}.
 \label{eq:gaussian}
 \end{equation}
%
We will compare its performance to the KDE likelihood for the Planck 2018 data in the two examples of section~\ref{sec:effective_likelihood} as well.

\subsection{Cumulative Optical Depth}

Although reionization PCs are mainly a tool for model testing, as they do not reconstruct the ionization history $x_e(z)$ itself, they do provide model-independent constraints on  the cumulative Thomson optical depth
\begin{equation}
\tau(z,z_{\rm max}) = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)},
\label{eq:cumtau}
\end{equation}
where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter.  The underlying reason is that CMB observables probe the optical depth and whether it is weighted toward high or low redshift through the amplitude and shape of 
$C_\ell^{EE}$ -- the higher the redshift, the
larger the relative contribution at higher $\ell$.

Specific models essentially place an implicit prior on the redshift dependence of the cumulative optical depth.  
%For example, the standard steplike model adopted in CAMB uses a tanh function to parameterize the hydrogen and singly ionized helium reionization
% \begin{equation}
%x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
% \label{eqn:tanh}
% \end{equation}
% where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$,
% excludes finite ionization above its transition redshift by definition.
The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no prior constraints on the mode amplitudes $m_a$ are imposed. The one subtlety when placing constraints on the cumulative optical depth is that the analysis also therfore allows
unphysical ionization fractions where $0<x_e<x_e^{\rm max}$  is not satisfied.
Therefore, when determining cumulative optical depth constraints as opposed to using reionization PCs as a tool for testing models,
 we do impose a physicality prior
 by truncating the posteriors of $m_a$
 {\it after} obtaining them from Markov Chain Monte Carlo sampling, following Ref.~\cite{Mortonson:2008rx}
%
\begin{equation}
\sum_{a=1}^5 m_a^2 \le (x_e^{\rm max}-x_e^{\rm fid})^2,
\end{equation}
where $x_e^{\rm fid}=0.15$ and $m_a^{-} \le m_a \le m_a^{+}$ with
\begin{equation}
m_a^{\pm} = \int_{z_{\rm min}}^{z_{\rm max} } dz \frac{S_a(z)[x_e^{\rm max} -2 x_e^{\rm fid}(z)]
\pm x_e^{\rm max} | S_a(z)|}{2(z_{\rm max}-z_{\rm min})}.
\label{eq:individualprior}
\end{equation}
%
Note that the original $0<x_e<x_e^{\rm max}$ condition cannot be strictly enforced here because we do not keep all but the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for physicality.  In other words, no physical model is excluded by these priors.
% I think we don't impose the prior at all
% for KDE 
%They are however noninformative priors when the PC constraints are being used for evaluating the likelihood of a series of physical models, which we will now describe in the next section.

%we employ the smoothed $x_e$ which formally has support beyond the bounds. 
% We include this small correction by integrating
% slightly past $z_{\rm max}$ in practice.  Here $n_{\rm H}(0)$ is the hydrogen number density at $z=0$.
   %Consequently we smooth the ionization history in Eq.~(\ref{eq:mmutoxe}) with a 
%Gaussian in $\mathrm{ln}(1+z)$ of width $\sigma_{\mathrm{ln}(1+z)} = 0.015$.

\wh{want a forward pointing sentence here about impact of priors in practice.}

\section{Planck 2018 PC Results}
\label{sec:results}
We now present the complete reionization constraints obtained from the Planck 2018 likelihoods using the principal components.
\subsection{Constraints on Principal Components}
We use the official Planck likelihoods [add citation] \texttt{plik\_lite} for the high-$\ell$ $TT$, $TE$ and $EE$ as well as \texttt{lowl} for the low-$\ell$ $TT$ throughout this paper. We have tested that our results do not change if in lieu of \texttt{plik\_lite} we used the full \texttt{plik\_full} likelihood in which the foreground parameters have not been marginalized over. For the low-$\ell$ $EE$ likelihood, we use the third-party released \texttt{srollv2} likelihood [add citation] in our official PC results and the effective likelihood code. In comparison with the official Planck-released \texttt{simall\_EE} likelihood, the \texttt{srollv2} likelihood had improved foreground cleaning, which enabled a factor of XX improvement in $\tau$ just in the steplike model [cite].

@@

\caption{Constraints from Planck data on the amplitudes of the five reionization principal components that describe all physical ionization models up to $z_{\rm max} = 30$. We show in red the Planck 2015 and the improved Planck 2018 contours in blue which uses the low-$\ell$ $EE$ likelihood \texttt{srollv2}. The 1D posterior distributions are shown on the diagonal and the 2D 68\% and 95\% C.L. regions in the $m_a-m_b$ planes in the lower triangle. The physicality priors on the ionization history used in this work are delineated by the box boundaries. [TODO: update legend] [Add text: The black solid line indicates the steplike models varying $\tau$. make point that tanh ML is now within the the center of PC constraints]}
\label{fig:plot_mjs_2018_vs_2015}
\end{figure*}
%results/pc_results/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite.pdf
%
In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the 1D posterior and 2D 68\% and 95\% confidence level contours for the 5 PC amplitudes that describe ionization models up to $\zmax=30$. The box boundaries correspond to the physicality priors shown in Eq.~\ref{eq:individualprior} and include all physical models along with some unphysical models. Results using the Planck 2018 data are shown in blue. While all five PC amplitudes are being constrained by Planck, the first two are particularly well-constrained. Contrary to the Planck 2015 results of Ref.~\cite{Heinrich:2016ojb} shown in red, the 2018 data prefers a much smaller amplitude for the first PC, leading to a reduced optical depth.
The trajectories shown in black lines are the steplike models of instantaneous reionization allowed by the Planck 2015 data with the arrow pointing in the direction of increasing optical depth. Clearly, these steplikes models are only a part of the allowed physical model space. By construction, they miss any high-redshift ionization component by assuming neglibile ionization levels before the transition redshift $z_{\rm re}$.

\begin{figure}
\includegraphics[width=0.40\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
\caption{A zoomed-in of Fig.~\ref{fig:plot_mjs_2018_vs_2015} on the best constrained $\zmax=30$ PC plane, the $m_1-m_2$ plane, comparing Planck 2015 (gray) vs Planck 2018 results using two different low-$\ell$ $EE$ likelihoods: the \texttt{simall\_EE} likelihood (red) vs the \texttt{srollv2} likelihood (blue), the latter had better foreground removal and is the one we show as our fiducial results throughout this paper.
}
\label{fig:plot_m1m2_2015_vs_2018}
\end{figure}

@@

We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, where we show beyond the physicality prior with the shaded area. [Discuss how the ellipse moved in the direction of lower tanh by going to lower m1.] 
In Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, we also show using the PC chain covariance $\bf C$ (black dashed lines) what corresponds to the simpler Gaussian likelihood of Eq.~\ref{eq:gaussian}. The exact values of the PC mean and $\bar{\bf m}$ and covariance $\bf C$ are shown in Table~\ref{tab:PC_stats}.

 This Gaussian approximation agrees well with the actual PC constraints, unlike in the Planck 2015 data where some differences were still visible on top of good agreement. The Gaussian likelihood is even faster than the kernel density estimate. As shall see in the next section for two explicit examples, the Gaussian likelihood shows generally good agreement with the full KDE result, but is unable to capture, as expected given its symmetric nature, any skewness in the PC distribution that may affect parameter constraints.

In Fig.~\ref{fig:plot_m1m2_lowE_vs_srollv2} we also compare the Planck 2018 results with two different low-$\ell$ $EE$ likelihoods: \texttt{simall\_EE} and \texttt{srollv2}. The \texttt{srollv2} contours are shrunk nearly directly along the lines of constant optical depth $\tau$ ($\tau_{12} = \tau_1 m_1 + \tau_2 m_2$) [wanna show this]? \wh{we might make this point in the context of Marius - that $\tau$ is mainly a linear function of $m_1$, $m_2$, and hence flat priors in $m's$ are locally flat in $\tau$}.

\section{Effective Likelihood}
\label{sec:effective_likelihood}

As described in ~\refsec{KDE}, we turn the PC constraints from the Planck 2018 into an effective likelihood code to constrain arbitrary models of ionization history between $6 \leq z \leq \zmax$. This code is available on GitHub at \url{} [give URL]. The initial release is offered for the Planck 2018 $\textsc{plik\_lite\_TTTTTE} + \textsc{lowl} + \textsc{srollv2}$ likelihood combination with $\zmax = 30$.

To obtain posteriors of model parameters from the corresponding Planck likelihoods, simply specify the functional form for $x_e(z)$ in the code, and run a MCMC analysis using the extension we provide to a MCMC sampler (currently supporting CosmoMC). Note that the likelihood cannot distinguish the effects of the model parameters that impact $z \sim 6$, where the Universe is assumed to be fully ionized. Furthermore, transitions faster than $\Delta z \sim 0.1$ between $6 < z < 6.25$ may not be fully captured because of how PC functions are discretized with $\delta z = 0.25$ and are linearly interpolated within this bin size. [review this statement] To avoid over-smoothing parameter posteriors while maintaining accuracy during the KDE operation, we suggest using the default value of $f = 0.14$. Finally, you may choose to run the chains at a different cosmological model than that of the default Planck 2018 best-fit model, provided it is close enough.

We demonstrate successful recovery of parameter posteriors using the effective likelihood code by comparing them to a direct MCMC analysis for two examples: 1) The tanh model, which is the standard approach used in CAMB; 2) a two-step toy model which has an one additional parameter to model the high-$z$ ionization as a plateau in $x_e$ before the transition to full ionization happens in the standard tanh model.

@@

The marginalized 1D constraints from the effective likelihood $\taulo = XX \pm XX$ and $\tauhi = XX \pm XX$ are consistent with those of the full likelihood $\taulo = XX \pm XX$ and $\tauhi = XX \pm XX$. The 95\% C.L. upper limit on high-redshift optical depth are also consistent: $\tau(15, 30) < XX$ for the effective likelihood and $\tau(15, 30) < XX$ for the full likelihood.

@@

\subsection{Cumulative optical depth constraints}

Let us now constrain the cumulative optical depth $\tau(z, \zmax)$ using the PC results. The optical depth is a better quantity than the ionization history $x_e$ for visualizing and comparing PC results because it is much more closely related to observables. Recall that these truncated PCs provide an accurate description of the models only in terms of their \textit{observable} impact, and are not meant to be used to reconstruct the exact ionization history.

>>>>

where we have discretized the redshift space with $\delta z= 0.25$, and where $\sigma_a^2$ are the variances of the PCs. We have rank-ordered the PCs from low to high variance, so that for the range $z_{\rm min} = 6$ (to be consistent with Ly$\alpha$ forest constraints, e.g.~\cite{Becker:2015lua}) and $z_{\rm max} = 30$, only the first 5 components are needed to describe all the information on $x_e$ carried by $C_\ell^{EE}$ to cosmic variance limit. In the work that follows, we therefore truncate the sum at $a = 1 .. n_{\rm{PC}}$, where $n_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ for the fiducial analysis. 
We also test its robustness by taking 
 for $z_{\rm max} = 50$ where $n_{\rm{PC}} = 7$ is required at the cosmic variance limit.

@@

For the Planck data set, most of the information on the ionization history comes from the first two modes which probe the amount of low vs.\ high redshift optical depth.   However all 5 PCs for $z_{\rm max}=30$ carry some constraint, with the higher modes controlling finer variations in redshift.

We follow Ref.~\cite{Heinrich:2018btc} to compute the CMB power spectra with PCs using a modified version of CAMB\footnote{CAMB: \url{http://camb.info}}~\cite{Lewis:1999bs, Howlett:2012mh}, where the PCs were discretized at $\delta z = 0.25$, and computed at the fiducial model $x_e^{\rm fid} = 0.15$. Note that following Ref.~\cite{Heinrich:2018btc}, we have updated the $\zmax = 30$ PCs to be computed with Planck 2015 rather than WMAP best-fit parameters, which results in minor differences between the PC functions used in Ref.~\cite{Heinrich:2016ojb} and here. For $z<6$, we follow CAMB and assume fully ionized hydrogen and singly ionized helium and for $z\leq 3.5$~\cite{Becker:2010cu}, doubly ionized helium~\cite{Becker:2010cu} with a width $\Delta z = 0.5$ (see also \S \ref{sec:example1}).

We obtain posterior constraints on the PC parameters $m_a$ from Markov Chain Monte Carlo sampling using COSMOMC\footnote{COSMOMC: \url{http://cosmologist.info/cosmomc}} of the relevant Planck likelihood discussed in \S \ref{sec:results} in an otherwise fiducial $\Lambda$CDM cosmology with the standard 5 parameters: the baryon density $\Omega_b h^2$, cold dark matter density 
$\Omega_c h^2$, effective angular sound horizon $\theta_{\rm MC}$, scalar power spectrum log amplitude $\ln (10^{10} A_s)$ and
its tilt $n_s$.  We take flat uninformative
priors in $m_a$ but discuss additional priors imposed by
removing demonstrably  unphysical ionization histories in \S \ref{sec:cumulative}.

\subsection{Effective PC Likelihood}
\label{sec:KDE}
The completeness property of the PCs enables us to turn PC chains obtained from a MCMC run into an effective likelihood that can be used to test any reionization model with the CMB without a cumbersome reanalysis at the level of power spectra data. In the following, we briefly recap the kernel density estimate technique used to build this likelihood and refer the readers to Ref.~\cite{Heinrich:2016ojb} for more details.
The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ for $i = 1$...$N_{\rm sample}$. Given any physical ionization history $x_e(z)$, we first obtain its PC representation $\mathbf{m}$ using Eq.~(\ref{eq:xetommu}). Since $\mathbf{m}$ can take any continuous value, we approximate its effective likelihood with a kernel density estimate
\beq
{\cal L}_{\rm PC}\left({\rm data}|\mathbf{m} \right)  = \sum_{i = 1}^{N_{\rm sample}} w_i K_f(\mathbf{m}-\mathbf{m}_i),
\eeq
where the overall normalization is arbitrary, and where we have chosen the smoothing kernel $K_f$ to be a  multivariate Gaussian with mean zero and covariance $f\mathbf{C}$, where $\mathbf{C}$ is the $N_{\rm PC} \times N_{\rm PC}$ covariance matrix estimated from the PC chains (see Table~\ref{tab:PC_stats}) and $f$ is a fraction smaller than 1.
For a Gaussian posterior, increasing the covariance by $1+f$ corresponds to increasing the standard deviation by approximately $1+f/2$. To minimize the amount of smoothing needed while still maintaining good accuracy in the tail of the distribution for any physical models, we oversample the PC distributions by running the PC chains far beyond convergence for about $N_{\sample} \approx 1.1 \times 10^{6}$ chain samples. 
% 1.06 cut w/ physicality prior
% 1.072 no cut
For this $N_{\sample}$, a fraction of $f = 0.14$ should be sufficient.
%Note also that we employ PC chains without the physicality priors allowing the smoothing kernel to cross over these priors. 
Evaluating the posterior distribution of model parameters $\bf p$ using the effective likelihood
\begin{equation}
P({\bf p}| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\bf p) \right] P(\bf p).
\end{equation}

This approach obviates the need to implement specific reionization models into CAMB and conduct separate COSMOMC sampling for each, thereby significantly reducing both human and computational effort.
For comparison, an even simpler effective likelihood is the Gaussian approximation of the $m_a$ posterior using the PC mean $\bar{\bf m}$ and covariance $\bf C$
%
\begin{equation}
 {\cal L}_{\rm gauss}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^{N_{\rm PC}} | \mathbf{C}| }}.
 \label{eq:gaussian}
 \end{equation}
%
In section~\ref{sec:effective_likelihood} we test the effective KDE and Gaussian approaches against a direct implementation
for two example models: the standard steplike model and a two-step model allowing for high-redshift ionization.

%
%
%
%We will compare its performance to the KDE likelihood for the Planck 2018 data in the two examples of section~\ref{sec:effective_likelihood} as well.

\subsection{Cumulative Optical Depth}
\label{sec:cumulative}

Although reionization PCs are mainly a tool for model testing using complete information from CMB polarization, as they do not reconstruct $x_e(z)$ itself, in particular the its rapidly  aspects, they do provide model-independent constraints on redshift integrated quantities such as  the cumulative Thomson optical depth
\begin{equation}
\tau(z,z_{\rm max}) = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)},
\label{eq:cumtau}
\end{equation}
where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter.  The underlying reason is that CMB observables probe the optical depth and whether it is weighted toward high or low redshift through the amplitude and shape of 
$C_\ell^{EE}$ -- the higher the redshift, the
larger the relative contribution at higher $\ell$.

Specific models essentially place an implicit prior on the redshift dependence of the cumulative optical depth.  
%For example, the standard steplike model adopted in CAMB uses a tanh function to parameterize the hydrogen and singly ionized helium reionization
% \begin{equation}
%x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
% \label{eqn:tanh}
% \end{equation}
% where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$,
% excludes finite ionization above its transition redshift by definition.
The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no range-bound prior constraints on the mode amplitudes $m_a$ are imposed. The one subtlety when placing constraints on the cumulative optical depth is that the analysis also allows
unphysical ionization fractions where $0 \le x_e \le x_e^{\rm max}$  is not satisfied.
Therefore, when determining cumulative optical depth constraints as opposed to using reionization PCs as a tool for testing models,
 we do impose a physicality prior
 by truncating the posteriors of $m_a$
 {\it after} obtaining them from Markov Chain Monte Carlo sampling, following Ref.~\cite{Mortonson:2008rx}
%
\begin{equation}
\sum_{a=1}^5 m_a^2 \le (x_e^{\rm max}-x_e^{\rm fid})^2,
\end{equation}
where $x_e^{\rm fid}=0.15$ and $m_a^{-} \le m_a \le m_a^{+}$ with
\begin{equation}
m_a^{\pm} = \int_{z_{\rm min}}^{z_{\rm max} } dz \frac{S_a(z)[x_e^{\rm max} -2 x_e^{\rm fid}(z)]
\pm x_e^{\rm max} | S_a(z)|}{2(z_{\rm max}-z_{\rm min})}.
\label{eq:individualprior}
\end{equation}
\wh{comment on the sum prior?  we don't seem to talk about it later - is it applied and is it always weaker in practice?}
\ch{I think it is always weaker, let me double check.}
%
Note that the original $0\le x_e \le x_e^{\rm max}$ condition cannot be strictly enforced here because we do not keep all but the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for physicality.  
In other words, no physical model is excluded by these priors but some unphysical models are included.  In \S \ref{sec:note_on_priors} we further explore the role of these and other choices of PC priors, and in \S\ref{sec:effective_likelihood} we test PC constraints against direct constraints for two example models.

\section{Planck 2018 PC Results}
\label{sec:results}
We now present the complete reionization constraints obtained from the Planck 2018 likelihoods using the principal components.
\subsection{Constraints on Principal Components}
We use the official Planck likelihoods [add citation] \texttt{plik\_lite} for the high-$\ell$ $TT$, $TE$ and $EE$ as well as \texttt{lowl} for the low-$\ell$ $TT$ throughout this paper. We have tested that our results do not change if in lieu of \texttt{plik\_lite} we used the full \texttt{plik\_full} likelihood in which the foreground parameters have not been marginalized over. For the low-$\ell$ $EE$ likelihood, we use the third-party released \texttt{srollv2} likelihood [add citation] in our official PC results and the effective likelihood code. In comparison with the official Planck-released \texttt{simall\_EE} likelihood, the \texttt{srollv2} likelihood improves foreground cleaning, 
which enables stronger reionization constraints, especially for the low multipole moments that are important for the low redshift constraints.

% can add a comment about impact 
%d a factor of XX improvement in $\tau$ just in the steplike model [cite].

@@

\caption{Constraints from Planck data on the amplitudes of the five reionization principal components that describe all physical ionization models up to $z_{\rm max} = 30$. We show in red the Planck 2015 and the improved Planck 2018 contours in blue which uses the low-$\ell$ $EE$ likelihood \texttt{srollv2}. The 1D posterior distributions are shown on the diagonal and the 2D 68\% and 95\% C.L. regions in the $m_a-m_b$ planes in the lower triangle. The physicality priors on the ionization history used in this work are delineated by the box boundaries. [TODO: update legend] [Add text: The black solid line indicates the steplike models varying $\tau$. make point that tanh ML is now within the the center of PC constraints]}
\label{fig:plot_mjs_2018_vs_2015}
\end{figure*}
%results/pc_results/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite.pdf
%
In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the 1D posterior and 2D 68\% and 95\% confidence level contours for the 5 PC amplitudes that describe ionization models up to $\zmax=30$. The shaded region in the $m_1-m_2$ plane represents the parameter region that would be excluded by an additional 
physicality prior from Eq.~(\ref{eq:individualprior}), whereas for the other cases the box itself represents this prior. Results using the Planck 2018 data are shown in blue. While all five PC amplitudes are  constrained by Planck, the first two are particularly well-constrained. Contrary to the Planck 2015 results of Ref.~\cite{Heinrich:2016ojb} shown in red, the 2018 data prefers a much smaller amplitude for the first PC, leading to a reduced optical depth.  \wh{This preference also increases the impact of placing a physicality prior at low values of $m_1$.
We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}.
In this excluded region, the optical depth of models are dominated by the low redshift end, but due to the fixed ionization at $z>z_{\rm min}=6$, the desired total optical depth cannot be achieved without an unphysical ionization elsewhere.} \ch{[Not sure if we should say ``In this excluded region" referring to the entire gray band, or specify overlap with data constraints where $m_2 < 0$ mostly?]}

In Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, we also show
the Planck 2018 constraints using the older \texttt{simall\_EE} likelihood.  The improvement 
from switching to the \texttt{srollv2} likelihood mainly strengthens the upper bound in the best constrained direction in the $m_1-m_2$ plane which shifts constraints toward higher \ch{[lower?]} total optical depth. \ch{[Or maybe you meant Planck 2015 to simall EE?]}

\wh{possibly point to a figure below with $\tau_{12}$ in}

%\wh{possibly move the discussion of trajectories into the model section}
%The trajectories shown in black lines are the steplike models of instantaneous reionization allowed by the Planck 2015 data with the arrow pointing in the direction of increasing optical depth. Clearly, these steplikes models are only a part of the allowed physical model space. By construction, they miss any high-redshift ionization component by assuming neglibile ionization levels before the transition redshift $z_{\rm re}$.

\begin{figure}
\includegraphics[width=0.40\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
\caption{A zoomed-in of Fig.~\ref{fig:plot_mjs_2018_vs_2015} on the best constrained $\zmax=30$ PC plane, the $m_1-m_2$ plane, comparing Planck 2015 (gray) vs Planck 2018 results using two different low-$\ell$ $EE$ likelihoods: the \texttt{simall\_EE} likelihood (red) vs the \texttt{srollv2} likelihood (blue), the latter had better foreground removal and is the one we show as our fiducial results throughout this paper. 
\wh{some of these figure versions look like low res bit maps - final versions should be vector graphics and look uniform in style between figures.} \ch{got it, will polish for sure.}
}
\label{fig:plot_m1m2_2015_vs_2018}
\end{figure}

@@

We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, where we show beyond the physicality prior with the shaded area. [Discuss how the ellipse moved in the direction of lower tanh by going to lower m1.]

In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we also show the constraints assuming a Gaussian posterior in $m_a$ using the PC chain covariance and means listed in Table~\ref{tab:PC_stats}.  The visual agreement with the full posteriors is remarkably good, especially in the well constrained components.  This suggests that the simple Gaussian  likelihood of Eq.~(\ref{eq:gaussian}) may suffice for rough constraints on models.    However we shall see that a quantitative comparison shows that there is enough skewness in the posteriors that the KDE likelihood 
still performs noticeably better.  \wh{revisit this 
statement once we firm up the comparison, e.g. what is noticeably better, means, 95\% confidence intervals, far tails?} \ch{yup, will do.}

% This Gaussian approximation agrees well with the actual PC constraints, unlike in the Planck 2015 data where some differences were still visible on top of good agreement. The Gaussian likelihood is even faster than the kernel density estimate. As shall see in the next section for two explicit examples, the Gaussian likelihood shows generally good agreement with the full KDE result, but is unable to capture, as expected given its symmetric nature, any skewness in the PC distribution that may affect parameter constraints.

%In Fig.~\ref{fig:plot_m1m2_lowE_vs_srollv2} we also compare the Planck 2018 results with two different low-$\ell$ $EE$ likelihoods: \texttt{simall\_EE} and \texttt{srollv2}. The \texttt{srollv2} contours are shrunk nearly directly along the lines of constant optical depth $\tau$ ($\tau_{12} = \tau_1 m_1 + \tau_2 m_2$) [wanna show this]? \wh{we might make this point in the context of Marius - that $\tau$ is mainly a linear function of $m_1$, $m_2$, and hence flat priors in $m's$ are locally flat in $\tau$}.

\section{Model Testing with PCs}
\label{sec:effective_likelihood}

We now turn to constraining reionization models with PCs using the
effective likelihood approach described in~\refsec{KDE}.  The benefit of this approach is that
complete constraints on any model of the ionization history  between $6 \leq z \leq \zmax$ can now be obtained without reanalyzing the Planck data. 
This effective likelihood code is available on GitHub at \url{} [give URL]. The initial release corresponds to our fiducial constraints derived from the Planck 2018 $\texttt{plik\_lite\_TTTTTE} + \texttt{lowl} + \texttt{srollv2}$ likelihood combination with $\zmax = 30$.

To use this approach, one simply specifies the functional form for $x_e(z)$ in the code, which then is converted to the PC parameters $m_a$ for the effective likelihood evaluation assuming a \ch{fiducial} $\Lambda$CDM cosmology.   
To obtain reionization model parameter posteriors, one samples this likelihood with any desired priors.   
We have also provided an MCMC sampler for models of high dimension (currently in CosmoMC). Note that the model parameters or priors must be arranged to explicitly satisfy 
full ionization below $z=6$. To avoid over-smoothing parameter posteriors while maintaining accuracy during the KDE operation, we suggest using the default value of $f = 0.14$.

Next we demonstrate the use and the successful recovery of parameter posteriors using the effective likelihood code by comparing them to a direct MCMC analysis for two examples: 1) The tanh model, which is the standard approach used in CAMB; 2) a two-step toy model which has one additional parameter modeling a high-$z$ ionization plateau before the transition to full ionization in the canonical tanh model.

@@

The marginalized 1D constraints from the effective likelihood $\taulo = XX \pm XX$ and $\tauhi = XX \pm XX$ are consistent with those of the full likelihood $\taulo = XX \pm XX$ and $\tauhi = XX \pm XX$. 
\wh{will be interesting to see if the numbers are more consistent with the $f$ smoothing than the visual plot is - that might help us see if getdist is responsible for extra smoothing that hides things.}
The 95\% C.L. upper limit on high-redshift optical depth are also consistent: $\tau(15, 30) < XX$ for the effective likelihood and $\tau(15, 30) < XX$ for the full likelihood.

@@

\subsection{Cumulative optical depth}

The differences between model dependent and model independent approaches are larger for the allowed cumulative optical depth $\tau(z, \zmax)$ at high $z$
compared to the total optical depth of the previous section.
The main results are again related to the $m_1$ and $m_2$ PC constraints. 
As discussed in Ref.~\cite{Heinrich:2016ojb}, raising $m_1$ increases the  optical depth at all $z$ but with a larger weighting at high $z$ whereas
raising $m_2$ shifts the optical depth from low $z$ to high $z$.  
Consequently, as shown in Fig.~\ref{fig:prior_box}, the second aspect of PC constraints beyond the well constrained total optical depth, is a positive
correlation between the two components that represents the high to low $z$ 
contribution to the total optical depth.  Any finer detail in the ionization history is much more poorly constrained.
